---
title: "Direct Circuitscape Runs from R"
author: "Bill Peterman"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: false
vignette: |
  %\VignetteIndexEntry{Direct Circuitscape Runs from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r libraries, message=FALSE, warning=FALSE}
library(ResistanceGA2)
library(terra)
```

# Overview

`Run_CS.ini()` is a low-level Circuitscape wrapper for cases where you want to
run Circuitscape directly from R without going through the optimization
workflow.

It complements `Run_CS.jl()` rather than replacing it.

- `Run_CS.jl()` is tailored to the package optimization workflow.
- `Run_CS.ini()` is tailored to direct Circuitscape runs.
- Both use Julia and `Circuitscape.jl` through `JuliaConnectoR`.

The main advantage of `Run_CS.ini()` is that it lets you work from R objects
while still thinking in terms of the Circuitscape `.ini` file.

# Julia Setup

You still need Julia and `Circuitscape.jl` installed.

```{r julia-home, eval=FALSE}
JULIA_HOME <- "C:/Users/your-name/AppData/Local/Programs/Julia-1.11.4/bin/"
Sys.setenv(JULIA_BINDIR = JULIA_HOME)
```

```{r julia-home-run, include=FALSE}
JULIA_HOME <- "C:/Users/peterman.73/AppData/Local/Programs/Julia-1.11.4/bin/"
Sys.setenv(JULIA_BINDIR = JULIA_HOME)
```

# Example Inputs

```{r example-inputs}
pts <- vect(samples[, 2:3], type = "points")
surface <- resistance_surfaces[["continuous"]]

pts
surface
```

# Pairwise Circuitscape with INI-Style Options

```{r run-cs-ini-pairwise, eval=FALSE}
pairwise_out <- Run_CS.ini(
  r = surface,
  CS_Point.File = pts,
  JULIA_HOME = JULIA_HOME,
  scenario = "pairwise",
  connect_four_neighbors_only = TRUE,
  write_cur_maps = TRUE,
  write_cum_cur_map_only = TRUE,
  return = "all",
  rm.files = FALSE
)
```

The returned object includes the raw Julia result, a parsed pairwise matrix
when one can be extracted, any current map written by Circuitscape, the final
config, and the paths to generated files.

```{r inspect-cs-ini, eval=FALSE}
names(pairwise_out)
pairwise_out$ini_file
pairwise_out$pairwise_matrix
pairwise_out$config
```

# Working from R Objects Instead of Manual Files

`Run_CS.ini()` can now write several Circuitscape support files directly from R
objects.

Supported convenience inputs include:

- `mask_file`
- `source_file`
- `ground_file`
- `included_pairs_file`
- `reclass_file`

## Advanced mode with source and ground rasters

```{r advanced-raster-mode, eval=FALSE}
source_r <- rast(surface)
ground_r <- rast(surface)
mask_r <- rast(surface)

source_vals <- rep(0, ncell(source_r))
ground_vals <- rep(0, ncell(ground_r))
source_vals[5] <- 1
ground_vals[ncell(ground_r) - 4] <- 1

values(source_r) <- source_vals
values(ground_r) <- ground_vals
values(mask_r) <- 1

advanced_out <- Run_CS.ini(
  r = surface,
  JULIA_HOME = JULIA_HOME,
  scenario = "advanced",
  source_file = source_r,
  ground_file = ground_r,
  ground_file_is_resistances = FALSE,
  mask_file = mask_r,
  return = "all",
  rm.files = FALSE
)
```

## Included pairs from an R table

```{r included-pairs, eval=FALSE}
keep_pairs <- data.frame(
  node1 = c(1, 1, 2),
  node2 = c(2, 3, 4)
)

pairwise_subset <- Run_CS.ini(
  r = surface,
  CS_Point.File = pts,
  JULIA_HOME = JULIA_HOME,
  included_pairs_file = keep_pairs,
  return = "all",
  rm.files = FALSE
)
```

Use exclude mode by passing a small list:

```{r exclude-pairs, eval=FALSE}
pairwise_exclude <- Run_CS.ini(
  r = surface,
  CS_Point.File = pts,
  JULIA_HOME = JULIA_HOME,
  included_pairs_file = list(
    mode = "exclude",
    pairs = keep_pairs
  ),
  return = "all"
)
```

## Reclassification tables from R

```{r reclass-table, eval=FALSE}
reclass_tbl <- data.frame(
  from = c(1, 2, 3),
  to = c(5, 25, 100)
)

reclass_out <- Run_CS.ini(
  r = resistance_surfaces[["categorical"]],
  CS_Point.File = pts,
  JULIA_HOME = JULIA_HOME,
  reclass_file = reclass_tbl,
  return = "all",
  rm.files = FALSE
)
```

Based on inspection of the current `Circuitscape.jl` source, I infer that
`reclass_file` remains part of the config template but may not yet be fully
acted on upstream. `Run_CS.ini()` still writes that file correctly from R.

# Existing Paths Still Work

If you already have files prepared on disk, you can still pass them directly.

```{r existing-paths, eval=FALSE}
Run_CS.ini(
  r = "my_surface.asc",
  CS_Point.File = "my_points.txt",
  JULIA_HOME = JULIA_HOME,
  mask_file = "mask.asc",
  source_file = "sources.asc",
  ground_file = "grounds.asc",
  included_pairs_file = "pairs.txt",
  return = "all"
)
```

# Inspecting the Generated INI

Keeping the files can be useful while debugging or comparing against a
hand-written Circuitscape setup.

```{r inspect-ini, eval=FALSE}
out <- Run_CS.ini(
  r = surface,
  CS_Point.File = pts,
  JULIA_HOME = JULIA_HOME,
  scenario = "pairwise",
  return = "all",
  rm.files = FALSE
)

readLines(out$ini_file)
```

# Return Modes

```{r return-modes, eval=FALSE}
raw_out <- Run_CS.ini(..., return = "compute")
matrix_out <- Run_CS.ini(..., return = "matrix")
map_out <- Run_CS.ini(..., return = "current_map")
all_out <- Run_CS.ini(..., return = "all")
```

- `compute` returns the raw Julia object
- `matrix` returns a parsed pairwise matrix when possible
- `current_map` returns the cumulative current map raster when written
- `all` returns the full bundle

# Relationship to the Optimization Workflow

If your goal is landscape genetic optimization, the usual workflow is still:

1. `GA.prep()`
2. `gdist.prep()` or `jl.prep()`
3. `SS_optim()`, `MS_optim()`, or `all_comb()`

Use `Run_CS.ini()` when you want more direct access to Circuitscape itself,
especially for advanced mode or for explicit `.ini`-style control.
