---
title: "Getting Started with ResistanceGA2"
author: "Bill Peterman"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Getting Started with ResistanceGA2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r libraries, message=FALSE, warning=FALSE}
library(ResistanceGA2)
library(terra)
```

# Overview

`ResistanceGA2` optimizes resistance surfaces with genetic algorithms.

- Resistance surfaces are supplied as `SpatRaster` objects.
- Sample locations are supplied as point `SpatVector` objects.
- Distance calculations can be done with `gdistance` or with Circuitscape in Julia.
- `gdistance` compatibility is handled internally; users do not need to create `raster` objects.
- The old `.asc` directory workflow, `sp` objects, standalone Circuitscape executable, and kernel-smoothing path are no longer part of the supported public API.

This vignette is a practical starting point for the current package workflow.

# Installation

```{r install, eval=FALSE}
install.packages("devtools")
devtools::install_github("wpeterman/ResistanceGA2", build_vignettes = TRUE)
```

If you want the package first and the documentation second, install without
building vignettes, then render the vignette sources later on a machine with
Pandoc available.

# Example Data

The package includes both larger optimization examples and lighter-weight
objects for vignette use.

```{r example-data}
pts_small <- vect(samples[, 2:3], type = "points")
pts_large <- vect(sample_pops[[1]], type = "points")

pts_small
resistance_surfaces
raster_orig
```

`samples` and `resistance_surfaces` are useful for lightweight examples and
precomputed outputs. `sample_pops`, `Dc_list`, and `raster_orig` are better
suited to full optimization runs.

# Step 1: Prepare GA Settings

`GA.prep()` expects a `SpatRaster` supplied through the `raster` argument.

For continuous surfaces, it is strongly recommended that transformation searches
stay within the Monomolecular families unless there is a strong biological or
ecological reason to expect a unimodal or quadratic-type relationship. Ricker
families are more flexible, but that flexibility can increase the risk of
overfitting when they are explored without a clear hypothesis.

```{r ga-prep, message=FALSE, warning=FALSE}
ga_inputs <- GA.prep(
  raster = resistance_surfaces[[c("categorical", "continuous")]],
  Results.dir = file.path(tempdir(), "ResistanceGA2-getting-started"),
  monitor = FALSE,
  quiet = TRUE
)

ga_inputs$layer.names
ga_inputs$surface.type
```

If you want to state that recommendation explicitly in your setup, pass
`select.trans = list("M", ...)` for continuous layers rather than opening the
search across all transformation families.

# Step 2: Choose a Distance Engine

## gdistance

Prepare a `gdistance` analysis by pairing a point `SpatVector` with the number
of populations and, when available, a response vector.

```{r gdist-prep, warning=FALSE}
gdist_inputs <- gdist.prep(
  n.Pops = nrow(samples),
  samples = pts_small,
  method = "costDistance"
)

names(gdist_inputs)
```

The package vignette data include a precomputed `gdistance` result built from
this exact setup.

```{r gdist-precomputed}
cd_matrix <- vignette_pairwise_outputs$gdistance_commute

dim(cd_matrix)
cd_matrix[1:4, 1:4]
```

To reproduce that matrix directly:

```{r gdist-live, eval=FALSE}
cd_matrix <- Run_gdistance(
  gdist.inputs = gdist_inputs,
  r = resistance_surfaces[["continuous"]]
)
```

## Circuitscape in Julia

If you want to use Circuitscape, install Julia and the `Circuitscape` Julia
package, then prepare inputs with `jl.prep()`. A Julia-specific guide is
provided in the separate `Julia_Guide` vignette.

`jl.prep()` and `Run_CS.jl()` need the Julia `bin/` directory. This will be specific to your local machine and must be uniquely specified.

```{r julia-home, eval=FALSE}
JULIA_HOME <- "C:/Users/your-name/AppData/Local/Programs/Julia-1.11.4/bin/"
Sys.setenv(JULIA_BINDIR = JULIA_HOME)
```

```{r julia-home-run, include=FALSE}
JULIA_HOME <- "C:/Users/peterman.73/AppData/Local/Programs/Julia-1.11.4/bin/"
Sys.setenv(JULIA_BINDIR = JULIA_HOME)
```

```{r jl-prep, eval=FALSE}
pairwise_genetic <- lower(as.matrix(dist(samples[, 2:3])))

jl_inputs <- jl.prep(
  n.Pops = nrow(samples),
  response = pairwise_genetic,
  CS_Point.File = pts_small,
  JULIA_HOME = Sys.getenv("JULIA_BINDIR"),
  run_test = TRUE
)
```

The package also ships a precomputed pairwise Circuitscape result and a
precomputed current map based on the same small example.

```{r jl-precomputed}
cs_vector <- vignette_pairwise_outputs$circuitscape_matrix

length(cs_vector)
head(cs_vector)
```

```{r current-map}
vignette_current_map
```

# Step 3: Fit Optimization Workflows

## Single-surface optimization

For a real optimization run, it is usually better to work from the larger toy
data bundled with the package.

```{r ss-workflow, eval=FALSE}
pts <- vect(sample_pops[[1]], type = "points")
gd <- lower(Dc_list[[1]])

GA.inputs <- GA.prep(
  raster = raster_orig,
  Results.dir = file.path(tempdir(), "ResistanceGA2-ss"),
  pop.size = 25,
  maxiter = 10,
  run = 5,
  monitor = FALSE,
  quiet = TRUE
)

gdist.inputs <- gdist.prep(
  n.Pops = nrow(sample_pops[[1]]),
  response = gd,
  samples = pts,
  method = "costDistance"
)

ss_out <- SS_optim(
  gdist.inputs = gdist.inputs,
  GA.inputs = GA.inputs,
  dist_mod = FALSE,
  null_mod = FALSE,
  diagnostic_plots = FALSE
)
```

## Multisurface optimization

```{r ms-workflow, eval=FALSE}
ms_out <- MS_optim(
  gdist.inputs = gdist.inputs,
  GA.inputs = GA.inputs,
  diagnostic_plots = FALSE
)
```

## All-combinations analysis

`all_comb()` requires `Results.dir = "all.comb"` at the `GA.prep()` stage, then
uses a separate results directory for the actual output. That output directory
will be created automatically when needed.

```{r all-comb, eval=FALSE}
GA.inputs_all <- GA.prep(
  raster = raster_orig,
  Results.dir = "all.comb",
  pop.size = 25,
  maxiter = 10,
  run = 5,
  monitor = FALSE,
  quiet = TRUE
)

results_dir <- file.path(tempdir(), "ResistanceGA2-all-comb")

all_out <- all_comb(
  gdist.inputs = gdist.inputs,
  GA.inputs = GA.inputs_all,
  results.dir = results_dir,
  max.combination = 2,
  iters = 10,
  replicate = 1
)
```

# Transformations and Inspection

Two useful helpers for inspecting resistance transformations are `Plot.trans()`
and `Resistance.tran()`.

```{r plot-trans}
Plot.trans(
  PARM = c(2.5, 100),
  Resistance = resistance_surfaces[["continuous"]],
  transformation = "Monomolecular",
  marginal.plot = FALSE
)
```

```{r resistance-tran}
transformed_surface <- Resistance.tran(
  transformation = "Monomolecular",
  shape = 2.5,
  max = 100,
  r = resistance_surfaces[["continuous"]]
)

transformed_surface
```

# Interpreting Outputs

The main optimization functions return lists containing:

- GA summaries and fitted models
- AICc tables for model comparison
- Pairwise distance objects used downstream
- saved plot and table outputs beneath the specified `Results.dir`

For larger or slower analyses, it is often worth running multiple replicates and
checking for consistency in the top-supported surfaces and parameter estimates.

# Migration Checklist

When updating older scripts, the key changes are:

- pass `SpatRaster` objects, not directories of `.asc` files
- pass point `SpatVector` objects, not `SpatialPoints`
- use `JuliaConnectoR` and Julia/Circuitscape.jl, not `JuliaCall`, `XRJulia`, or the standalone `.exe`
- do not rely on the removed `scale = TRUE` / kernel-smoothing workflow

If you stay within that contract, the package workflow is much simpler and much
more consistent than the old mixed raster/file/path interface.
