---
title: "Joint Isolation by Environment and Isolation by Resistance"
author: "Bill Peterman"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: false
vignette: |
  %\VignetteIndexEntry{Joint Isolation by Environment and Isolation by Resistance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r libraries, message=FALSE, warning=FALSE}
library(ResistanceGA2)
library(terra)
```

```{r saved-output-flags, include=FALSE}
output_dir <- system.file("extdata", "ibe-ibr-demo", package = "ResistanceGA2")

if (!nzchar(output_dir)) {
  local_extdata <- file.path("inst", "extdata", "ibe-ibr-demo")
  if (dir.exists(local_extdata)) {
    output_dir <- local_extdata
  }
}

have_saved_outputs <- file.exists(file.path(output_dir, "analysis_summary.rds"))

if (have_saved_outputs) {
  summary_bundle <- readRDS(file.path(output_dir, "analysis_summary.rds"))
  ss_out <- readRDS(file.path(output_dir, "single_surface_optim.rds"))
  ms_out <- readRDS(file.path(output_dir, "multi_surface_optim.rds"))
}
```

# Overview

This vignette sketches a practical workflow for fitting isolation by
environment (`IBE`) and isolation by resistance (`IBR`) concurrently with the
new MLPE interface in `ResistanceGA2`.

The central idea is simple:

- `IBE` enters the model as one or more pairwise endpoint-difference
  covariates.
- `IBR` enters the model as the optimized resistance-distance term.
- both are estimated in the same MLPE framework so we can ask whether one still
  explains genetic differentiation after controlling for the other.

This example uses the package's Idaho demo data:

- `rga2_demo` for sites, extracted site environments, and the chord-distance matrix
- `rga2_demo_covariates` for the raster layers used to generate resistance distances

The analysis is structured so that the expensive optimization steps can be run
outside the vignette, saved as package extdata, and then displayed here for
later exploration.

The document is organized in two layers:

- a separate builder script that generates the full analysis objects and output files
- this vignette, which then loads those saved results by default for display
  and interpretation

The optimization examples below use an explicit seed so the worked example is
reproducible. The external builder script also supports named runtime profiles
and an optional core-count override so you can choose between quick smoke tests
and deeper optimization runs without editing the script itself.

```{r output-note}
if (!have_saved_outputs) {
  cat(
    "Saved analysis outputs were not found under ",
    output_dir,
    ".\n\nRun `data-raw/build_ibe_ibr_vignette_outputs.R` first if you want ",
    "the full set of model tables, optimization summaries, and plots to be ",
    "stored beneath `inst/extdata/ibe-ibr-demo/` and appear throughout this ",
    "vignette.\n",
    sep = ""
  )
}
```

# Design Choices

For a joint `IBE`/`IBR` analysis, it is important not to blur the mechanism of
endpoint mismatch with the mechanism of landscape permeability. In this example:

- `IBE` is represented by site-level mismatch in `cti` and `hli`
- `IBR` is optimized over the separate raster hypotheses `err27` and `ffp`
- geographic distance is included as an additional covariate

This is only one defensible partition. The key point is to avoid using the same
variable in both roles unless that dual interpretation is explicitly part of
the biological hypothesis.

# Demo Data

```{r demo-data}
data(rga2_demo)
data(rga2_demo_covariates)

if (inherits(rga2_demo_covariates, "PackedSpatRaster")) {
  rga2_demo_covariates <- unwrap(rga2_demo_covariates)
}

r_covariate <- subset(rga2_demo_covariates, c("err27", "ffp", "cti", "hli"))

sites <- rga2_demo$sites
site_env <- rga2_demo$site_environment

sites[, c("demo_id", "SiteName", "x", "y")]
site_env
r_covariate
```

# Step 1: Build Pairwise Geographic and Environmental Covariates

The pairwise response is the lower triangle of the 12 x 12 chord-distance
matrix. We pair that response with:

- Euclidean distance among sites (`IBD`)
- pairwise endpoint differences in `cti` and `hli` (`IBE`)

There are two common ways to encode `IBE`, and they answer slightly different
questions.

Absolute environmental differences:

- use one pairwise covariate per environmental variable, such as
  `|cti_i - cti_j|` or `|hli_i - hli_j|`
- keep interpretation simple, because each coefficient refers to a specific
  environmental mismatch axis
- are the best starting point when you have a small set of clear, named
  hypotheses

Multivariate environmental distance:

- compress several site-level variables into a single pairwise distance term
- is useful when you want an overall environmental dissimilarity index rather
  than separate coefficients
- is harder to interpret mechanistically, because one coefficient summarizes
  several variables at once

In practice, I would usually start with absolute differences for a small number
of biologically motivated variables, then compare that model to a simpler
multivariate-distance version as a sensitivity check.

That strategy is especially useful here because `cti` and `hli` have
interpretable ecological meanings, while `env_dist` is intentionally more
agnostic. Separate absolute-difference terms tell us which environmental axis
is carrying signal; a multivariate distance tells us whether there is overall
environmental mismatch even when we do not want to interpret individual axes
too literally.

```{r pairwise-covariates}
site_ids <- sites$demo_id
pair_index <- t(combn(site_ids, 2))
pair_index <- data.frame(
  from = pair_index[, 1],
  to = pair_index[, 2],
  stringsAsFactors = FALSE
)

coords <- rga2_demo$sample_coords
geo <- as.numeric(scale(lower(as.matrix(dist(coords)))))

ibe_diff <- site_pairwise_diff(
  site_env[, c("cti", "hli")],
  scale = TRUE
)

env_dist <- site_pairwise_dist(
  site_env[, c("cti", "hli")],
  scale = TRUE
)

head(ibe_diff)
head(env_dist)
```

`ibe_diff` and `env_dist` are both valid `IBE` representations, but they should
not usually be included in the same model because they encode overlapping
information.

# Step 1a: Check Predictor Overlap Before Fitting Models

Before interpreting coefficients, it helps to inspect how strongly the pairwise
predictors overlap. If `IBD`, `IBE`, and later `IBR` are all highly correlated,
then model ranking may still be informative but individual fixed-effect
estimates will be harder to interpret mechanistically.

```{r predictor-correlation}
predictor_cor <- cor(
  data.frame(
    geo = geo,
    cti_diff = ibe_diff$cti_diff,
    hli_diff = ibe_diff$hli_diff,
    env_dist = env_dist$env_dist
  )
)

round(predictor_cor, 3)
```

If you have already run the external builder script, the saved output bundle
stores this same matrix for later review:

```{r predictor-correlation-saved}
if (have_saved_outputs) {
  round(summary_bundle$predictor_correlation, 3)
}
```

# Step 2: Fit Baseline MLPE Models Before Optimization

Before optimizing any resistance surfaces, it is useful to ask whether there is
support for `IBD`, `IBE`, or both.

```{r baseline-pair-data}
pair_data <- mlpe_data(
  response = rga2_demo$genetic_vector,
  pairs = pair_index,
  geo = geo,
  cti_diff = ibe_diff$cti_diff,
  hli_diff = ibe_diff$hli_diff,
  env_dist = env_dist$env_dist
)

head(pair_data)
```

# Optional: Multiple Dyadic Random Effects

The examples above use one MLPE dyad term, `(1 | pair)`, where `pair` is built
from the `from` and `to` site columns. The current `mlpe()` implementation can
also include **more than one separate dyadic random-intercept term** in the same
model, as long as each term is linked to its own pair of endpoint columns
through the `pairs` argument.

This can be useful when pairwise observations share endpoints at multiple
sampling scales. For example, we may want one dyadic random effect for repeated
use of the same sampled sites and a second dyadic random effect for repeated use
of broader region labels attached to those sites.

One current constraint is that each dyadic term still expects the two endpoint
labels to differ within every row for that term. A coarse region-level dyadic
effect therefore works naturally for **between-region** pairs, but not for
within-region rows where `from_region == to_region`. The example below keeps
only between-region observations so the two-term fit is well-defined.

```{r multiple-dyad-terms}
site_region <- setNames(
  ifelse(sites$x <= stats::median(sites$x), "west", "east"),
  as.character(sites$demo_id)
)

pair_data_region <- pair_data
pair_data_region$from_region <- site_region[as.character(pair_data_region$from)]
pair_data_region$to_region <- site_region[as.character(pair_data_region$to)]
pair_data_region <- pair_data_region[
  pair_data_region$from_region != pair_data_region$to_region,
  ,
  drop = FALSE
]
rownames(pair_data_region) <- NULL

m_site_region <- mlpe(
  response ~ geo + cti_diff + hli_diff +
    (1 | site_pair) + (1 | region_pair),
  data = pair_data_region,
  pairs = list(
    site_pair = c("from", "to"),
    region_pair = c("from_region", "to_region")
  )
)

summary(m_site_region)
```

The important distinction is that these are **parallel dyadic terms**, not a
nested dyadic syntax. Formula shorthand such as `(1 | site_pair / region_pair)`
is not currently translated into a nested MLPE covariance structure, and
dyadic terms with same-label endpoints are still rejected. If you need both
site-level and region-level dyadic effects right now, write them as separate
terms and ensure the retained rows satisfy that endpoint-label constraint.

```{r baseline-models, eval=FALSE}
m_null <- mlpe(
  response ~ 1 + (1 | pair),
  data = pair_data,
  pairs = c("from", "to")
)

m_ibd <- mlpe(
  response ~ geo + (1 | pair),
  data = pair_data,
  pairs = c("from", "to")
)

m_ibe_cti <- mlpe(
  response ~ cti_diff + (1 | pair),
  data = pair_data,
  pairs = c("from", "to")
)

m_ibe_hli <- mlpe(
  response ~ hli_diff + (1 | pair),
  data = pair_data,
  pairs = c("from", "to")
)

m_ibe_both <- mlpe(
  response ~ cti_diff + hli_diff + (1 | pair),
  data = pair_data,
  pairs = c("from", "to")
)

m_ibd_ibe <- mlpe(
  response ~ geo + cti_diff + hli_diff + (1 | pair),
  data = pair_data,
  pairs = c("from", "to")
)

m_ibd_envdist <- mlpe(
  response ~ geo + env_dist + (1 | pair),
  data = pair_data,
  pairs = c("from", "to")
)

baseline_table <- data.frame(
  model = c(
    "Null",
    "IBD",
    "IBE: cti",
    "IBE: hli",
    "IBE: cti + hli",
    "IBD + IBE",
    "IBD + env_dist"
  ),
  AIC = c(
    AIC(m_null),
    AIC(m_ibd),
    AIC(m_ibe_cti),
    AIC(m_ibe_hli),
    AIC(m_ibe_both),
    AIC(m_ibd_ibe),
    AIC(m_ibd_envdist)
  ),
  logLik = c(
    logLik(m_null),
    logLik(m_ibd),
    logLik(m_ibe_cti),
    logLik(m_ibe_hli),
    logLik(m_ibe_both),
    logLik(m_ibd_ibe),
    logLik(m_ibd_envdist)
  )
)

baseline_table[order(baseline_table$AIC), ]
```

This baseline comparison tells us whether the environmental mismatch terms have
signal before we add the path-based `IBR` component.

The different baseline models answer different questions:

- `Null` tells us how much structure is captured by the MLPE random effect alone
- `IBD` asks whether simple geographic separation explains genetic distance
- `IBE: cti` and `IBE: hli` isolate each environmental mismatch axis
- `IBE: cti + hli` asks whether a small set of interpretable mismatch variables
  improves fit together
- `IBD + IBE` tests whether environmental mismatch adds anything after
  accounting for geography
- `IBD + env_dist` is a compact alternative when you prefer one overall
  environmental dissimilarity term

If saved outputs are available, the table below shows an actual model ranking
from the example analysis.

```{r baseline-models-saved}
if (have_saved_outputs) {
  summary_bundle$baseline_table
}
```

This fixed-effect table is often even more informative than the rank table,
because it shows how much the environmental coefficients shrink once geography
enters the model.

```{r baseline-fixef-saved}
if (have_saved_outputs) {
  summary_bundle$baseline_fixef
}
```

For this example, the fast-run saved outputs suggest a simple but instructive
pattern:

- `IBD` is clearly supported relative to `Null`
- `IBE` by itself is weaker than `IBD`
- adding `cti_diff` and `hli_diff` to `IBD` changes the coefficients only
  modestly in the fast run, suggesting that most of the baseline signal is
  geographic rather than environmental
- the multivariate `env_dist` model behaves similarly, which makes it a useful
  sensitivity check rather than a replacement for the more interpretable
  univariate terms

# Step 3: Prepare the Conditional IBR Workflow

Once the fixed covariates are chosen, they can be supplied directly to
`gdist.prep()` so every resistance optimization is fit conditionally on those
same `IBD` and `IBE` terms.

```{r gdistance-inputs}
pts <- vect(
  rga2_demo$sample_coords,
  type = "points",
  crs = crs(r_covariate, proj = TRUE)
)

workflow_covariates <- data.frame(
  geo = geo,
  cti_diff = ibe_diff$cti_diff,
  hli_diff = ibe_diff$hli_diff
)

gdist_inputs <- gdist.prep(
  n.Pops = nrow(rga2_demo$sample_coords),
  response = rga2_demo$genetic_vector,
  samples = pts,
  covariates = workflow_covariates,
  formula = gd ~ geo + cti_diff + hli_diff,
  method = "costDistance"
)

gdist_inputs$formula
head(gdist_inputs$df)
```

The resulting optimization models are of the form:

```r
gd ~ geo + cti_diff + hli_diff + cd + (MLPE random effects)
```

where `cd` is the resistance-distance term generated from the candidate raster(s).

This conditional formulation is the key to the joint framework. We are not
asking whether `IBR` is better than `IBE` in isolation; we are asking whether a
path-based resistance hypothesis still improves model fit after geography and
endpoint environmental mismatch have already been accounted for.

# Step 4: Run Single-Surface IBR Models Conditional on IBE

This example treats `err27` and `ffp` as the candidate permeability hypotheses.

```{r single-surface-setup, eval=FALSE}
ibr_layers <- subset(r_covariate, c("err27", "ffp"))

GA_inputs_ss <- GA.prep(
  raster = ibr_layers,
  Results.dir = file.path(tempdir(), "rga2-ibe-ibr-single"),
  method = "LL",
  seed = 123,
  pop.size = 25,
  maxiter = 15,
  run = 5, 
  parallel = F, 
  keepBest = FALSE,
  monitor = FALSE,
  quiet = TRUE
)

ss_out <- SS_optim(
  gdist.inputs = gdist_inputs,
  GA.inputs = GA_inputs_ss,
  dist_mod = TRUE,
  null_mod = TRUE,
  diagnostic_plots = FALSE
)

ss_out$AICc
summary(ss_out)
```

That table compares:

- the optimized `IBR + IBD + IBE` model for each candidate raster
- a pure geographic `Distance` model
- a `Null` model with no fixed predictors beyond the intercept

If saved outputs are available, we can inspect the actual optimization ranking:

```{r ss-saved-table}
if (have_saved_outputs) {
  summary_bundle$ss_aicc
}
```

The single-surface summary is also helpful because it packages the model
ranking with a concise narrative about the best-supported surface.

```{r ss-saved-summary}
if (have_saved_outputs) {
  summary(ss_out)
}
```

And the corresponding plot provides a quick visual ranking of the candidate
surfaces:

```{r ss-saved-plot, fig.width=7, fig.height=5}
if (have_saved_outputs) {
  plot(ss_out)
}
```

The exact top-ranked surface can vary across systems, and with shallow search
settings it can also vary across exploratory runs if no seed is set. The key
interpretive question is whether at least one candidate resistance surface
outranks the pure `Distance` model while `IBD + IBE` are already in the fixed
effects. If it does, that is evidence for an additional path-based landscape
effect.

# Step 5: Run a Multi-Surface IBR Model Conditional on IBE

If the single-surface results suggest complementary hypotheses, a multisurface
fit lets us ask whether a composite `IBR` term improves support further.

```{r multi-surface, eval=FALSE}
GA_inputs_ms <- GA.prep(
  raster = ibr_layers,
  Results.dir = file.path(tempdir(), "rga2-ibe-ibr-multi"),
  method = "LL",
  seed = 123,
  pop.size = 25,
  maxiter = 15,
  run = 5,
  parallel = FALSE,
  keepBest = FALSE,
  monitor = FALSE,
  quiet = TRUE
)

ms_out <- MS_optim(
  gdist.inputs = gdist_inputs,
  GA.inputs = GA_inputs_ms,
  diagnostic_plots = FALSE
)

ms_out$AICc.tab
ms_out$percent.contribution
summary(ms_out)
```

Multisurface optimization is especially helpful when two rasters are plausible
for different ecological reasons. It lets us ask whether the best explanation
is truly composite or whether one layer dominates once they are optimized
together.

```{r ms-saved-summary}
if (have_saved_outputs) {
  summary(ms_out)
}
```

```{r ms-saved-plot, fig.width=7, fig.height=5}
if (have_saved_outputs) {
  plot(ms_out)
}
```

The percent-contribution table is often more informative than the raw
multisurface fit by itself. It tells us whether two plausible resistance
hypotheses are both contributing meaningfully or whether one layer dominates
once they are optimized together. That is scientifically useful because it
shows that not every plausible raster necessarily adds unique explanatory power
in the joint model.

# Step 6: Generate Full Analysis Outputs Outside the Vignette

The intended workflow is to run a separate builder script once, then let this
vignette pull those saved outputs back in for a fully populated worked example.

The package includes a builder script that runs the baseline MLPE models plus
single- and multi-surface optimization and saves the results as `.rds` and
`.csv` files beneath `inst/extdata/ibe-ibr-demo/` by default so the completed
worked example can ship with the package.

Run from the package root:

```{r build-script, eval=FALSE}
source("data-raw/build_ibe_ibr_vignette_outputs.R")
build_ibe_ibr_demo_outputs()
```

From an RStudio session, you can specify the parameters directly:

```r
source("data-raw/build_ibe_ibr_vignette_outputs.R")

build_ibe_ibr_demo_outputs(
  profile = "default",
  cores = NULL,
  seed = 123
)
```

For a quick smoke test:

```r
source("data-raw/build_ibe_ibr_vignette_outputs.R")

build_ibe_ibr_demo_outputs(
  profile = "fast",
  cores = NULL,
  seed = 123
)
```

For a deeper run:

```r
source("data-raw/build_ibe_ibr_vignette_outputs.R")

build_ibe_ibr_demo_outputs(
  profile = "full",
  cores = NULL,
  seed = 123
)
```

The function arguments are:

- `profile = "fast"` for quick smoke tests
- `profile = "default"` for a moderate reproducible example
- `profile = "full"` for a deeper optimization run
- `cores` as an optional positive integer passed to `GA.prep(parallel = ...)`
- `seed` for reproducible GA initialization

The `cores` argument is now passed directly through to `GA.prep(parallel = ...)`,
including on Windows. For large optimization runs, this is the main knob for
scaling the workflow across available hardware.

The command-line interface still works if you prefer `Rscript`, but the
function-based RStudio workflow is the recommended path for interactive use.

The script writes:

- `pair_data.rds`
- `workflow_covariates.rds`
- `baseline_models.rds`
- `baseline_model_table.csv`
- `baseline_fixed_effects.csv`
- `single_surface/Results/` with optimizer summaries, rasters, and distance matrices
- `single_surface_optim.rds`
- `single_surface_aicc.csv`
- `multi_surface/Results/` with multisurface summaries and percent contribution
- `multi_surface_optim.rds`
- `multi_surface_aicc.csv`
- `analysis_summary.rds`

# Step 7: Load and Explore Saved Outputs

```{r load-saved, eval=FALSE}
output_dir <- system.file("extdata", "ibe-ibr-demo", package = "ResistanceGA2")

baseline_models <- readRDS(file.path(output_dir, "baseline_models.rds"))
ss_out <- readRDS(file.path(output_dir, "single_surface_optim.rds"))
ms_out <- readRDS(file.path(output_dir, "multi_surface_optim.rds"))
summary_bundle <- readRDS(file.path(output_dir, "analysis_summary.rds"))

summary_bundle$baseline_table
summary_bundle$ss_aicc
summary_bundle$ms_aicc

plot(ss_out)
plot(ms_out)
```

# Interpretation

This framework lets us separate three related questions:

1. Does genetic differentiation increase with geographic distance alone?
2. Does endpoint environmental mismatch explain additional variation?
3. Does a resistance-distance hypothesis still improve fit after controlling
   for both geography and environment?

Possible outcomes are:

- `IBE` improves fit but `IBR` does not: endpoint mismatch matters more than
  permeability.
- `IBR` improves fit after conditioning on `IBE`: there is support for a
  genuine landscape-connectivity effect.
- both improve fit: connectivity and environmental mismatch likely contribute
  independently.
- coefficients become unstable or change sign across models: the predictors are
  probably too correlated for strong mechanistic interpretation.

The example here also shows why it is useful to explore several model classes
instead of going straight to optimization:

- baseline MLPE models tell us whether `IBE` is promising enough to include at all
- conditional single-surface optimization tells us whether specific `IBR`
  hypotheses improve fit beyond `IBD + IBE`
- multisurface optimization tells us whether apparently useful resistance
  layers still contribute once they compete in the same model

That sequence turns the vignette into more than a software tutorial. It becomes
an analysis template for asking:

1. What is the best-supported baseline explanation of pairwise genetic structure?
2. Does environmental mismatch add anything to geography?
3. Does landscape permeability add anything after both of those are already in the model?
4. If multiple resistance hypotheses are plausible, do they all matter, or does one dominate?

The same workflow can be extended by:

- replacing `cti` and `hli` with other site-level environmental variables
- adding a single multivariate `env_dist` term instead of several univariate
  mismatch terms
- moving from `gdistance` to `jl.prep()` if Circuitscape-based pair filtering
  or current maps are required
