---
title: "Using Julia with ResistanceGA2"
author: "Bill Peterman"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Using Julia with ResistanceGA2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r libraries, message=FALSE, warning=FALSE}
library(ResistanceGA2)
library(terra)
```

# Overview

`ResistanceGA2` supports Circuitscape through Julia and `JuliaConnectoR`.

- The standalone Circuitscape executable is no longer part of the supported workflow.
- `JuliaCall` and `XRJulia` are not used.
- Users still supply `terra` objects directly; temporary text, raster, and `.ini`
  files are written internally only as needed for Circuitscape.

This guide focuses on installing Julia, verifying the connection from R, and
running the package’s Julia-backed workflow.

# Install Julia and Circuitscape.jl

1. Install Julia from <https://julialang.org/downloads/>.
2. Open Julia and install Circuitscape:

```julia
using Pkg
Pkg.add("Circuitscape")
Pkg.test("Circuitscape")
```

3. Install `ResistanceGA2` from GitHub if needed:

```{r install, eval=FALSE}
install.packages("devtools")
devtools::install_github("wpeterman/ResistanceGA2", build_vignettes = TRUE)
```

# Point R to Julia

`jl.prep()` and `Run_CS.jl()` need the Julia `bin/` directory. This will be specific to your local machine and must be uniquely specified.

```{r julia-home, eval=FALSE}
JULIA_HOME <- "C:/Users/your-name/AppData/Local/Programs/Julia-1.11.4/bin/"
Sys.setenv(JULIA_BINDIR = JULIA_HOME)
```

```{r julia-home-run, include=FALSE}
JULIA_HOME <- "C:/Users/peterman.73/AppData/Local/Programs/Julia-1.11.4/bin/"
Sys.setenv(JULIA_BINDIR = JULIA_HOME)
```

After R knows where Julia lives, `JuliaConnectoR` will start Julia lazily on
the first `juliaEval()` or `juliaCall()` invocation. In other words, the calls
below do not test an already-open connection; they are what trigger the initial
Julia startup and then verify that evaluation works.

```{r verify-julia, eval=TRUE}
Sys.setenv(JULIA_BINDIR = JULIA_HOME)
JuliaConnectoR::juliaEval("string(VERSION)")
JuliaConnectoR::juliaEval("using Circuitscape")
```

# Prepare a Small Julia Example

The package includes a lightweight example that is convenient for setup checks.

```{r small-example}
pts <- vect(samples[, 2:3], type = "points")
surface <- resistance_surfaces[["continuous"]]
pairwise_genetic <- lower(as.matrix(dist(samples[, 2:3])))

pts
surface
```

Now create the Julia inputs. Setting `run_test = TRUE` is a useful smoke test
the first time you connect Julia to the package on a machine.

```{r jl-prep, eval=FALSE}
jl_inputs <- jl.prep(
  n.Pops = nrow(samples),
  response = pairwise_genetic,
  CS_Point.File = pts,
  JULIA_HOME = JULIA_HOME,
  run_test = TRUE,
  silent = TRUE
)
```

# Run Circuitscape from R

## Pairwise resistance distances

To compute pairwise effective resistance distances directly:

```{r run-circuitscape, eval=FALSE}
cs_vector <- Run_CS.jl(
  jl.inputs = jl_inputs,
  r = surface,
  output = "matrix"
)
```

The package ships a precomputed example result from this same lightweight
workflow:

```{r precomputed-cs}
length(vignette_pairwise_outputs$circuitscape_matrix)
head(vignette_pairwise_outputs$circuitscape_matrix)
```

## Direct Circuitscape runs with `.ini`-style control

If you want fuller control over Circuitscape options than `Run_CS.jl()` is
designed to expose, use `Run_CS.ini()`.

```{r run-cs-ini, eval=FALSE}
cs_out <- Run_CS.ini(
  r = surface,
  CS_Point.File = pts,
  JULIA_HOME = JULIA_HOME,
  scenario = "pairwise",
  connect_four_neighbors_only = TRUE,
  write_cur_maps = TRUE,
  write_cum_cur_map_only = TRUE,
  return = "all",
  rm.files = FALSE
)
```

See the separate `Direct Circuitscape Runs from R` vignette for examples using
advanced mode, source and ground inputs, included-pairs files, and other
Circuitscape support files.

## Current maps

To compute a cumulative current map, set `CurrentMap = TRUE` and request raster
output. `EXPORT.dir` can be any directory path; the wrapper handles path
normalization internally.

```{r run-current-map, eval=FALSE}
current_map <- Run_CS.jl(
  jl.inputs = jl_inputs,
  r = surface,
  CurrentMap = TRUE,
  output = "raster",
  EXPORT.dir = file.path(tempdir(), "ResistanceGA2-current-map")
)
```

The package also includes a precomputed current map from that same example:

```{r precomputed-current-map}
vignette_current_map
```

# Use Julia in Optimization

Once `jl_inputs` is ready, the optimization workflow looks just like the
`gdistance` workflow except that you pass `jl.inputs` instead of `gdist.inputs`.
The same transformation guidance applies here: for continuous surfaces, it is
usually best to restrict searches to Monomolecular families unless you have a
strong biological or ecological reason to test unimodal or quadratic-type
Ricker relationships.

```{r julia-optim, eval=FALSE}
GA.inputs <- GA.prep(
  raster = resistance_surfaces[["continuous"]],
  Results.dir = file.path(tempdir(), "ResistanceGA2-julia"),
  pop.size = 20,
  maxiter = 10,
  run = 5,
  monitor = FALSE,
  quiet = TRUE
)

ss_out <- SS_optim(
  jl.inputs = jl_inputs,
  GA.inputs = GA.inputs,
  dist_mod = FALSE,
  null_mod = FALSE,
  diagnostic_plots = FALSE
)
```

These optimization outputs are now S3 objects, so routine inspection is
cleaner than before.

```{r julia-s3-ss, eval=FALSE}
class(ss_out)
summary(ss_out)
plot(ss_out)
```

For multiple surfaces:

```{r julia-ms, eval=FALSE}
GA.inputs_multi <- GA.prep(
  raster = resistance_surfaces[[c("categorical", "continuous")]],
  Results.dir = file.path(tempdir(), "ResistanceGA2-julia-multi"),
  pop.size = 20,
  maxiter = 10,
  run = 5,
  monitor = FALSE,
  quiet = TRUE
)

ms_out <- MS_optim(
  jl.inputs = jl_inputs,
  GA.inputs = GA.inputs_multi,
  diagnostic_plots = FALSE
)
```

```{r julia-s3-ms, eval=FALSE}
class(ms_out)
summary(ms_out)
plot(ms_out)
```

# Tuning and Troubleshooting

Some options are especially useful when working with Julia-backed runs:

- `run_test = TRUE` in `jl.prep()` verifies the Julia/Circuitscape connection before a long run.
- `cholmod = TRUE` is usually fastest, but can be memory-intensive on large rasters.
- `cholmod = FALSE, precision = TRUE` switches to the single-precision path when memory is limiting.
- `scratch` is helpful if the default temporary directory is slow or restricted.
- `rm.files = FALSE` can be useful while debugging generated Circuitscape inputs and outputs.
- `write.files` and `write.criteria` let you keep the generated `.ini` and raster files for selected runs.

If Julia is installed but the package cannot connect:

- confirm `JULIA_HOME` or `JULIA_BINDIR` points to Julia’s `bin/` directory
- run `JuliaConnectoR::juliaEval("using Circuitscape")` manually
- make sure `Circuitscape.jl` is installed in the Julia environment you are actually invoking

# Summary

The Julia backend is now the supported Circuitscape path for `ResistanceGA2`.
Once Julia and `Circuitscape.jl` are working, the rest of the package workflow
remains close to the `gdistance` workflow: create `GA.inputs`, create
`jl.inputs`, then run `SS_optim()`, `MS_optim()`, `all_comb()`, or direct
`Run_CS.jl()` calls as needed.
