---
title: "Using Julia and Circuitscape with ResistanceGA2"
author: "Bill Peterman"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: false
vignette: |
  %\VignetteIndexEntry{Using Julia and Circuitscape with ResistanceGA2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  message   = FALSE,
  warning   = FALSE
)
```

```{r libraries}
library(ResistanceGA2)
library(terra)
```

---

# Overview

`ResistanceGA2` supports two distance engines: `gdistance` (which runs entirely inside R) and Circuitscape running through Julia (which requires an external Julia installation). This vignette covers everything needed to install, configure, and troubleshoot the Julia/Circuitscape path.

**When to use Julia/Circuitscape vs. gdistance:**

| Criterion | Julia/Circuitscape | gdistance |
|---|---|---|
| Output matches Circuitscape | Yes | No |
| Current maps | Yes | No |
| Speed (iterative optimization) | Slower per call (startup overhead) | Faster per call |
| External dependency | Julia + Circuitscape.jl | None |
| Random-walk distances | Yes (resistance distance) | Yes (commuteDistance) |
| Least-cost path distances | No | Yes (costDistance) |

For most iterative optimization workflows, `gdistance::commuteDistance()` is the most convenient choice as it stays inside R, requires no external setup, and produces qualitatively similar distances to Circuitscape resistance distance. Use Julia/Circuitscape when you need to:

- Produce output that directly matches a Circuitscape workflow (for comparability with published analyses).
- Generate cumulative current maps for visualization or downstream analysis.
- Use Circuitscape's cholmod-accelerated solver for very large landscapes.

---

# Step 1: Install Julia

## Downloading Julia

Download Julia from the official website: <https://julialang.org/downloads/>

Install the **current stable release** (e.g., Julia 1.11.x). Both 64-bit Windows and macOS/Linux installers are available. The default installation location works well; note the installation path because you will need it later.

**Default installation paths by platform:**

- **Windows:** `C:\Users\<username>\AppData\Local\Programs\Julia-X.Y.Z\`
- **macOS:** `/Applications/Julia-X.Y.app/Contents/Resources/julia/`
- **Linux:** Wherever you extracted the tarball, typically `/opt/julia-X.Y.Z/`

The Julia binary directory (`bin/`) is one level inside the installation root. For example, on Windows: `C:\Users\<username>\AppData\Local\Programs\Julia-1.11.4\bin\`

## Verifying the Julia Installation

Open a terminal (Command Prompt, PowerShell, or bash) and run:

```bash
julia --version
```

You should see output like `julia version 1.11.4`. If the command is not found, Julia's `bin/` directory is not on your PATH. You may need to add it manually, or just supply the full path when configuring `ResistanceGA2`.

---

# Step 2: Install Circuitscape.jl

Open Julia (type `julia` at the terminal prompt, or open the Julia application) and install the Circuitscape package:

```julia
using Pkg
Pkg.add("Circuitscape")
Pkg.test("Circuitscape")
```

The installation will download Circuitscape and its dependencies. The test suite confirms the installation is working correctly. This step only needs to be done once per Julia environment.

> **Note:** Julia has a package manager that maintains environments similar to R's `renv` or Python's `venv`. By default, packages you install from the Julia REPL go into the default environment (`@v1.11` for Julia 1.11). When `ResistanceGA2` launches Julia via `JuliaConnectoR`, it uses this same default environment. If you later switch Julia versions, you will need to install Circuitscape again in the new version's default environment.

## Installing JuliaConnectoR

Install the `JuliaConnectoR` R package, which provides the bridge between R and Julia:

```{r install-juliaconnector, eval=FALSE}
install.packages("JuliaConnectoR")
```

---

# Step 3: Configure R to Find Julia

`ResistanceGA2` needs to know where Julia's `bin/` directory is. Set this before calling any Julia-related functions:

```{r julia-home, eval=FALSE}
# Replace with your actual Julia bin/ path
JULIA_HOME <- "C:/Users/your-name/AppData/Local/Programs/Julia-1.11.4/bin/"
Sys.setenv(JULIA_BINDIR = JULIA_HOME)
```

```{r julia-home-run, include=FALSE}
JULIA_HOME <- "C:/Users/peterman.73/AppData/Local/Programs/Julia-1.11.4/bin/"
Sys.setenv(JULIA_BINDIR = JULIA_HOME)
```

To avoid setting this every session, add the `Sys.setenv()` line to your `.Rprofile`:

```{r rprofile, eval=FALSE}
# Add to ~/.Rprofile (or project-level .Rprofile):
Sys.setenv(JULIA_BINDIR = "C:/Users/your-name/AppData/Local/Programs/Julia-1.11.4/bin/")
```

> **Note:** On Windows, use forward slashes (`/`) in path strings in R, even though Windows normally uses backslashes. `JuliaConnectoR` and `ResistanceGA2` normalize paths internally, but using forward slashes avoids escape-sequence issues.

## Finding Julia's Path Automatically

If Julia is on your system PATH, you can find the `bin/` directory from R:

```{r find-julia, eval=FALSE}
# Find Julia on PATH
julia_exe <- Sys.which("julia")
if (nzchar(julia_exe)) {
  JULIA_HOME <- dirname(julia_exe)
  message("Found Julia at: ", JULIA_HOME)
} else {
  message("Julia not found on PATH. Set JULIA_BINDIR manually.")
}
```

---

# Step 4: Verify the Connection

After setting `JULIA_BINDIR`, verify that R can start Julia and load Circuitscape:

```{r verify-julia, eval=TRUE}
Sys.setenv(JULIA_BINDIR = JULIA_HOME)

# This triggers Julia startup on first call
JuliaConnectoR::juliaEval("string(VERSION)")
```

```{r verify-circuitscape, eval=TRUE}
# Confirm Circuitscape is available
JuliaConnectoR::juliaEval("using Circuitscape")
```

If both commands succeed without error, your Julia/Circuitscape setup is ready for `ResistanceGA2`.

---

# Step 5: Prepare a Small Julia Example

## Set Up Sample Data

```{r small-example}
pts     <- vect(samples[, 2:3], type = "points")
surfaces <- terra::unwrap(resistance_surfaces)
surface  <- terra::subset(surfaces, "continuous")

pts
surface
```

## Create Julia Inputs with jl.prep()

`jl.prep()` configures everything needed for Circuitscape runs. Setting `run_test = TRUE` is strongly recommended the first time you use the package on a new machine; it runs a small Circuitscape test to verify the full pipeline before a long optimization:

```{r jl-prep, eval=FALSE}
pairwise_genetic <- lower(as.matrix(dist(samples[, 2:3])))

jl_inputs <- jl.prep(
  n.Pops        = nrow(samples),
  response      = pairwise_genetic,
  CS_Point.File = pts,
  JULIA_HOME    = JULIA_HOME,
  run_test      = TRUE,   # verify the connection
  silent        = TRUE    # suppress Julia startup messages
)

names(jl_inputs)
```

### jl.prep() Parameter Details

| Parameter | Default | Description |
|---|---|---|
| `n.Pops` | — | Number of sample populations |
| `response` | NULL | Pairwise genetic distances (vector from `lower()`) |
| `CS_Point.File` | — | `SpatVector` of sample point locations |
| `JULIA_HOME` | — | Path to Julia `bin/` directory |
| `run_test` | FALSE | Run a test Circuitscape calculation to verify setup |
| `silent` | FALSE | Suppress Julia startup messages |
| `cholmod` | TRUE | Use CHOLMOD direct solver (faster, more memory) |
| `precision` | FALSE | Use single precision (saves memory for large landscapes) |
| `scratch` | NULL | Temporary file directory (use if default tmpdir is slow) |
| `rm.files` | TRUE | Remove intermediate files after each run |

**Solver selection guidance:**

- `cholmod = TRUE` (default): Uses the CHOLMOD Cholesky factorization. Fastest option but can require substantial memory for large rasters. Start here.
- `cholmod = FALSE` (iterative): Falls back to iterative methods. Slower but uses less peak memory. Try this if CHOLMOD runs out of memory.
- `precision = TRUE`: Uses single-precision arithmetic, roughly halving memory use. Slightly less accurate but acceptable for most landscape genetics analyses.

---

# Computing Pairwise Resistance Distances

## Run_CS.jl() for Pairwise Distances

In this workflow, `Run_CS.jl(..., output = "matrix")` returns a lower-triangle
vector of pairwise resistance distances, ordered the same way as other
`ResistanceGA2` pairwise outputs. That is why the result is indexed like a
vector, not like a 2-D matrix.

```{r run-circuitscape, eval=FALSE}
cs_vector <- Run_CS.jl(
  jl.inputs = jl_inputs,
  r         = surface,
  output    = "matrix"
)

length(cs_vector)
head(cs_vector)
```

The precomputed example for the small dataset ships with the package:

```{r precomputed-cs}
cs_vector <- vignette_pairwise_outputs$circuitscape_matrix

length(cs_vector)
head(cs_vector)
```

## Run_CS.jl() Parameter Details

| Parameter | Default | Description |
|---|---|---|
| `jl.inputs` | — | Output of `jl.prep()` |
| `r` | — | `SpatRaster` resistance surface |
| `CurrentMap` | FALSE | Compute cumulative current map in addition to pairwise distances |
| `output` | `"matrix"` | `"matrix"` for pairwise distances; `"raster"` for current map |
| `EXPORT.dir` | NULL | Directory for output files when `output = "raster"` |
| `rm.files` | TRUE | Remove temporary files after run |
| `write.files` | FALSE | Write output files even when temporary |
| `write.criteria` | NULL | Only write files when LL exceeds this threshold |

---

# Computing Current Maps

Current maps show the cumulative flow of "current" across the entire landscape between all pairs of source nodes. They are useful for identifying high-connectivity corridors, bottlenecks, and pinch points.

```{r run-current-map, eval=FALSE}
export_dir <- file.path(tempdir(), "ResistanceGA2-current-map")

current_map <- Run_CS.jl(
  jl.inputs  = jl_inputs,
  r          = surface,
  CurrentMap = TRUE,
  output     = "raster",
  EXPORT.dir = export_dir
)

plot(current_map, main = "Cumulative current map")
```

The package ships a precomputed current map for the small example:

```{r precomputed-current-map}
plot(vignette_current_map)
```

---

# Julia in Optimization Workflows

Once `jl_inputs` is ready, the optimization workflow is identical to the `gdistance` workflow: simply pass `jl.inputs` instead of `gdist.inputs`:

```{r julia-optim, eval=FALSE}
GA.inputs <- GA.prep(
  raster      = terra::subset(terra::unwrap(resistance_surfaces), "continuous"),
  Results.dir = file.path(tempdir(), "ResistanceGA2-julia"),
  pop.size    = 25,
  maxiter     = 200,
  run         = 25,
  monitor     = FALSE,
  quiet       = TRUE
)

ss_out <- SS_optim(
  jl.inputs        = jl_inputs,
  GA.inputs        = GA.inputs,
  dist_mod         = FALSE,
  null_mod         = FALSE,
  diagnostic_plots = FALSE
)
```

For multiple surfaces:

```{r julia-ms, eval=FALSE}
GA.inputs_multi <- GA.prep(
  raster      = terra::subset(
    terra::unwrap(resistance_surfaces),
    c("categorical", "continuous")
  ),
  Results.dir = file.path(tempdir(), "ResistanceGA2-julia-multi"),
  pop.size    = 25,
  maxiter     = 200,
  run         = 25,
  monitor     = FALSE,
  quiet       = TRUE
)

ms_out <- MS_optim(
  jl.inputs        = jl_inputs,
  GA.inputs        = GA.inputs_multi,
  diagnostic_plots = FALSE
)
```

> **Tip:** During optimization, `ResistanceGA2` calls `Run_CS.jl()` hundreds or thousands of times (once per candidate solution per generation). Julia's startup overhead is paid only once when `JuliaConnectoR` first connects; subsequent calls are much faster. However, file I/O for the intermediate Circuitscape inputs and outputs still occurs each call. Setting `rm.files = TRUE` (the default) keeps the temporary directory clean and avoids accumulating large numbers of files.

---

# Comparing gdistance and Julia Outputs

## Conceptual Similarity

Both `gdistance::commuteDistance()` and Circuitscape compute random-walk based effective distances. They are derived from the same mathematical framework (circuit theory / random walk theory) and should produce highly correlated outputs on the same resistance surface.

In practice, the outputs are not identical because:

1. `commuteDistance()` uses a graph-based representation where only adjacent cells are connected, while Circuitscape models current flow through the full grid using finite element methods.
2. `commuteDistance()` uses 4-neighbor or 8-neighbor connectivity; Circuitscape uses 4-neighbor by default.
3. Numerical precision and solver differences introduce small discrepancies.

The two distance measures will typically be very highly correlated (r > 0.95) for the same resistance surface, but the absolute values will differ in scale.

## Comparing Outputs

```{r compare-outputs, eval=FALSE}
# Compute both distance types from the same surface
gdist_inputs <- gdist.prep(
  n.Pops = nrow(samples),
  samples = terra::vect(samples[, 2:3], type = "points"),
  method = "commuteDistance"
)

gdist_out <- Run_gdistance(
  gdist.inputs = gdist_inputs,
  r            = terra::subset(terra::unwrap(resistance_surfaces), "continuous")
)

# These will be on different scales but should be highly correlated
gdist_vec <- lower(as.matrix(gdist_out))
cs_vec    <- vignette_pairwise_outputs$circuitscape_matrix

cor(gdist_vec, cs_vec)

# Visual comparison
plot(gdist_vec, cs_vec,
     xlab = "gdistance commuteDistance",
     ylab = "Circuitscape resistance distance",
     main = "Comparison of distance engines")
```

The high correlation means that optimization results (which surface is best, what parameters are recovered) are typically very similar between the two engines. **The choice of distance engine should be driven by what you need from the output** (Circuitscape-matched values, current maps, etc.), not by expectation of meaningfully different optimization results.

---

# Performance Considerations

## Landscape Size

Circuitscape computational cost scales approximately with the number of non-null cells in the resistance raster. Larger landscapes take longer.

| Raster size (cells) | Approximate runtime per call | Notes |
|---|---|---|
| < 50,000 | < 1 second | Negligible overhead |
| 50,000 – 500,000 | 1–30 seconds | Optimization may be slow |
| > 500,000 | 30+ seconds | Consider aggregating raster |
| > 2,000,000 | Minutes per call | Impractical for iterative optimization |

For large landscapes, consider:

1. **Aggregating the raster** to a coarser resolution for optimization, then resampling the optimal surface back to fine resolution for current map generation.
2. **Using `gdistance`** for the iterative optimization step, then running Circuitscape once with the optimal parameters for final output.
3. **Using `precision = TRUE`** to reduce memory use.
4. **Using `cholmod = FALSE`** if CHOLMOD runs out of memory.

## Number of Populations

Circuitscape runtime also scales with the number of source nodes (sample populations). For n populations, there are n*(n-1)/2 pairwise computations.

| n populations | Pairwise observations | Typical overhead per call |
|---|---|---|
| 10 | 45 | Minimal |
| 20 | 190 | Moderate |
| 50 | 1225 | Substantial |
| 100 | 4950 | Consider gdistance instead |

For datasets with many populations on large landscapes, `gdistance` will usually be substantially faster for optimization. Use Circuitscape for the final output only.

---

# Environment Management for Julia Packages

Julia uses a built-in package manager (Pkg) that stores package versions in environment manifests. The key environment for `ResistanceGA2` is the **default environment** for the Julia version you're using.

## Checking Your Julia Environment

From the Julia REPL:

```julia
using Pkg
Pkg.status()
```

This shows all packages installed in the current environment. You should see `Circuitscape` listed.

## Multiple Julia Versions

If you have multiple Julia versions installed, make sure `JULIA_HOME` points to the one where you installed `Circuitscape.jl`. Different Julia versions have separate environments.

```{r multiple-julia, eval=FALSE}
# Check which Julia version JuliaConnectoR is using
JuliaConnectoR::juliaEval("string(VERSION)")

# If wrong, update JULIA_HOME and restart R
Sys.setenv(JULIA_BINDIR = "/path/to/correct/julia/bin/")
```

## Updating Circuitscape.jl

To update Circuitscape to the latest version, run from the Julia REPL:

```julia
using Pkg
Pkg.update("Circuitscape")
```

After updating, test the new version before running a large optimization:

```{r test-after-update, eval=FALSE}
jl_inputs_test <- jl.prep(
  n.Pops        = nrow(samples),
  CS_Point.File = vect(samples[, 2:3], type = "points"),
  JULIA_HOME    = JULIA_HOME,
  run_test      = TRUE
)
```

---

# Troubleshooting

## Julia Not Found

**Symptom:** Error: "Could not find Julia" or "JULIA_BINDIR not set"

**Solutions:**
1. Verify `JULIA_BINDIR` is set correctly: `Sys.getenv("JULIA_BINDIR")`
2. Confirm the path exists and contains the Julia executable:
   ```{r check-julia-path, eval=FALSE}
   bindir <- Sys.getenv("JULIA_BINDIR")
   dir.exists(bindir)
   file.exists(file.path(bindir, "julia.exe"))   # Windows
   file.exists(file.path(bindir, "julia"))       # macOS/Linux
   ```
3. Set the path explicitly before calling any package functions:
   ```{r set-julia, eval=FALSE}
   Sys.setenv(JULIA_BINDIR = "C:/path/to/julia/bin/")
   ```

## Circuitscape Not Found

**Symptom:** `using Circuitscape` fails with "Package not installed"

**Solutions:**
1. Open Julia directly (not from R) and run `Pkg.add("Circuitscape")`.
2. Confirm you are using the correct Julia version: `JuliaConnectoR::juliaEval("string(VERSION)")`.
3. Check that Circuitscape was installed in the default environment, not a project-specific environment:
   ```julia
   using Pkg
   Pkg.activate()        # activate default environment
   Pkg.add("Circuitscape")
   ```

## Slow Startup

**Symptom:** The first call to `jl.prep()` or `Run_CS.jl()` takes 30–60 seconds.

This is normal Julia behavior (JIT compilation on first use). Subsequent calls within the same R session will be much faster because Julia caches compiled code. The startup cost is paid only once per R session.

To minimize the impact on optimization runs, call `jl.prep()` once at the beginning of your session (which triggers Julia startup) before beginning optimization.

## Out of Memory

**Symptom:** Circuitscape crashes or returns an error about insufficient memory.

**Solutions:**
1. Use `precision = TRUE` in `jl.prep()` to switch to single-precision arithmetic.
2. Use `cholmod = FALSE` to switch from the direct CHOLMOD solver to an iterative solver.
3. Aggregate the resistance raster to a coarser resolution.
4. Reduce the number of sample populations if feasible.

## File Permission Errors on Windows

**Symptom:** Error messages about inability to write to temporary files.

**Solutions:**
1. Set `scratch` in `jl.prep()` to a directory where you have write permissions:
   ```{r scratch-example, eval=FALSE}
   jl_inputs <- jl.prep(
     ...,
     scratch = "C:/Users/your-name/julia_scratch/"
   )
   ```
2. Verify that the default `tempdir()` is writable: `dir.exists(tempdir())`

## Connection Drops Mid-Optimization

**Symptom:** Julia-related errors appear after the optimization has been running for a while.

This can occur if Julia's memory or disk use causes the process to be killed. Solutions:

1. Monitor system resources during the run.
2. Use `precision = TRUE` and/or `cholmod = FALSE` to reduce memory use.
3. Aggregate the raster to a coarser resolution.
4. Split the optimization into smaller chunks (fewer surfaces, smaller GA population).

---

# Summary

The Julia backend provides Circuitscape-compatible pairwise resistance distances and cumulative current maps from within R. Once Julia and Circuitscape.jl are installed and `JULIA_BINDIR` is set, the interface is nearly identical to the `gdistance` workflow:

```
jl.prep()       →   jl_inputs
Run_CS.jl()     →   pairwise distances or current map
SS_optim()      →   (pass jl.inputs instead of gdist.inputs)
MS_optim()      →   (pass jl.inputs instead of gdist.inputs)
all_comb()      →   (pass jl.inputs instead of gdist.inputs)
```

For iterative optimization on large landscapes or large sample sets, consider using `gdistance::commuteDistance()` for the optimization step and running Circuitscape once with the optimal parameters for final output.

---

# References

McRae, B. H. 2006. Isolation by resistance. *Evolution* 60:1551–1561.

McRae, B. H., B. G. Dickson, T. H. Keitt, and V. B. Shah. 2008. Using circuit theory to model connectivity in ecology, evolution, and conservation. *Ecology* 89:2712–2724.

Peterman, W. E. 2018. ResistanceGA: An R package for the optimization of resistance surfaces using genetic algorithms. *Methods in Ecology and Evolution* 9:1638–1647.

Schtickzelle, N., M. Baguette, and H. van Dyck. 2022. Individual variation in dispersal drives population dynamics and genetic structure. *Nature Ecology and Evolution* (early view).

Shah, V. B., and B. H. McRae. 2008. Circuitscape: A tool for landscape ecology. Proceedings of the 7th Python in Science Conference (SciPy 2008).
