---
title: "ResistanceGA2: Terra-Based Workflows"
author: "Bill Peterman"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{ResistanceGA2: Terra-Based Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r libraries, message=FALSE, warning=FALSE}
library(ResistanceGA2)
library(terra)
```

# Background

`ResistanceGA2` is the terra-based continuation of the original
`ResistanceGA` framework for optimizing resistance surfaces with genetic
algorithms. The core idea is unchanged from the approach first laid out in
Peterman et al. (2014) and later formalized for the package in Peterman
(2018): rather than assigning resistance values entirely from expert opinion
or searching only a small grid of candidate values, the package uses
stochastic optimization to search a much broader parameter space and identify
surfaces that best explain pairwise genetic differentiation.

That framework is useful for continuous surfaces, categorical surfaces, and
multiple-surface hypotheses. Continuous layers can be transformed through
parameterized response curves, categorical layers can have class-specific
resistance values estimated, and multisurface models can combine several
hypothesized drivers into a single composite surface. The current package keeps
that modeling framework but modernizes the implementation so that the package 
uses `terra`, Circuitscape is run through Julia with `JuliaConnectoR`, and
`gdistance` remains available through internal raster conversion.

Resistance distances can now be calculated in two main ways. `gdistance`
supports least-cost path distances with `costDistance()` and random-walk
commute-time distances with `commuteDistance()`. Julia/Circuitscape provides
pairwise resistance distances and current maps. In practice, `gdistance` is
often the most convenient starting point for iterative optimization because it
stays inside R and is usually faster for repeated model fitting, while
Julia/Circuitscape remains essential when you need direct Circuitscape output
or cumulative current maps.

This is still a model-based and computationally intensive workflow. Runtime
depends on raster size, number of sample locations, number of surfaces,
transformation space, and optimizer settings. You must think carefully about 
biological realism, raster resolution, and sampling design before launching 
large search spaces. If you plan to interpret resistance distances or current 
flow biologically, it is worth also reading the broader resistance-modeling and
circuit-theory literature rather than treating the package as a black box.

The practical interface in `ResistanceGA2` is simpler than the legacy package.

- Resistance surfaces should be supplied as `SpatRaster` objects.
- Sample locations should be supplied as point `SpatVector` objects.
- `gdistance` is still supported, but raster conversion happens internally.
- Circuitscape is supported through Julia and `JuliaConnectoR`, not through the standalone `.exe`.
- The old kernel-smoothing / `kernel.scale` workflow has been removed.

This vignette focuses on the current package framework rather than the legacy
file-based workflow.

# Thoughts on Optimization

There are three objective functions available in `GA.prep()`: `"AIC"`,
`"R2"`, and `"LL"`. The default in `ResistanceGA2` is still log-likelihood
(`method = "LL"`), and that is usually the best place to start. `R2` is
intuitive, but in practice it is often not sensitive enough to provide a clean
objective surface for the genetic algorithm. Recent simulations have also called into question the use of R-squared (Beninde et al. 2024, Molec Eco Res). AIC- and AICc-based criteria can
still be useful for comparing fitted models, but they depend on assumptions
about effective sample size and model complexity that are not fully
straightforward for pairwise MLPE models.

That ambiguity matters most when deciding how to count `k`, the number of
parameters used in AIC/AICc calculations. The fitted MLPE model always contains
a single fixed-effect resistance term, but the resistance distances themselves
may have been produced by optimizing several underlying parameters. Continuous
surfaces estimate a transformation family plus shape and maximum resistance;
categorical surfaces estimate multiple class-specific values; multisurface
models combine information from several layers into a single composite
predictor. Because of that, there is no universally accepted way to count `k`
in this setting.

`ResistanceGA2` keeps that uncertainty explicit by allowing `k.value` to
control how model complexity is counted. Different defensible choices can lead
to different AIC/AICc penalties, so users should report which option they used
and why. The default `k.value = 2` is a practical starting point, but it
should not be treated as uniquely correct for every study design.

A useful general strategy is to begin with simpler, biologically plausible
hypotheses and only expand the search space when the data and system justify
it. Repeated runs should recover broadly similar results if the signal is
strong. It is also highly recommended that continuous transformations be
restricted to the Monomolecular families unless there is a strong biological or
ecological reason to expect a unimodal or quadratic-type relationship. Ricker
transformations are more flexible, but that flexibility also increases the risk
of overfitting.

When choosing between distance engines, `gdistance::commuteDistance()` is often
a good default for iterative model fitting because it captures the same
random-walk style logic as Circuitscape resistance distance while remaining
fast and easy to run inside R. Julia/Circuitscape is still valuable when you
need direct Circuitscape outputs, want to produce current maps, or need to
match an existing Circuitscape-based workflow more closely. In either case, it
is best to treat optimization as one step in a broader inference process rather
than as a fully automated truth-finding procedure.

# Example Data

The package includes example resistance surfaces, sample locations, and genetic-distance matrices.

```{r example-data}
pts <- vect(sample_pops[[1]], type = "points")
gd <- lower(Dc_list[[1]])
pts_small <- vect(samples[, 2:3], type = "points")

pts
pts_small
raster_orig
```

`sample_pops` is stored as coordinate matrices for convenience, but the current workflow converts those coordinates to `SpatVector` objects before analysis.

For the expensive distance-engine examples below, this vignette uses packaged,
precomputed outputs generated from the smaller `samples` point set and the
`resistance_surfaces[["continuous"]]` layer. That keeps the vignette fast to
render and avoids requiring Julia during documentation builds. The builder
script used to refresh those objects lives in `data-raw/build_vignette_outputs.R`.

# Prepare Optimization Inputs

## Genetic Algorithm Settings

`GA.prep()` now expects a `SpatRaster` supplied through the `raster` argument.

For continuous surfaces, the recommended default is to restrict transformation
searches to the Monomolecular families. Ricker families should usually only be
explored when there is a strong biological or ecological basis for a unimodal
or quadratic-type relationship; otherwise, their added flexibility can increase
the risk of overfitting.

```{r ga-prep, message=FALSE, warning=FALSE}
ga_inputs <- GA.prep(
  raster = raster_orig,
  Results.dir = file.path(tempdir(), "ResistanceGA2-vignette"),
  monitor = FALSE,
  quiet = TRUE
)

ga_inputs$layer.names
ga_inputs$surface.type
```

In practice, this means many analyses should keep `select.trans` at its default
Monomolecular-only behavior or specify `list("M", ...)` explicitly for
continuous layers.

## Pairwise Distances with gdistance

Use `gdist.prep()` when you want to optimize against least-cost or commute-time distances.

```{r gdist-prep, warning=FALSE}
gdist_demo_inputs <- gdist.prep(
  n.Pops = nrow(samples),
  samples = pts_small,
  method = "costDistance"
)

names(gdist_demo_inputs)
```

You can also calculate a distance matrix directly from a resistance surface.
The matrix shown here is precomputed and shipped with the package.

```{r run-gdistance}
cd_matrix <- vignette_pairwise_outputs$gdistance_commute

dim(cd_matrix)
```

To reproduce that object manually:

```{r run-gdistance-live, eval=FALSE}
cd_matrix <- Run_gdistance(
  gdist.inputs = gdist_demo_inputs,
  r = resistance_surfaces[["continuous"]]
)
```

## Pairwise Distances with Circuitscape in Julia

Use `jl.prep()` when you want to run Circuitscape through Julia. This requires a working Julia installation plus the `Circuitscape` Julia package.

```{r jl-prep, eval=FALSE}
jl_inputs <- jl.prep(
  n.Pops = nrow(samples),
  CS_Point.File = pts_small,
  JULIA_HOME = Sys.getenv("JULIA_BINDIR"),
  run_test = FALSE
)
```

The packaged vignette outputs also include a precomputed pairwise Circuitscape
result generated from the same surface and sample locations.

```{r jl-precomputed}
cs_result <- vignette_pairwise_outputs$circuitscape_matrix

length(cs_result)
head(cs_result)
```

# Optimization Workflows

## Single-Surface Optimization

```{r ss-optim, eval=FALSE}
ss_out <- SS_optim(
  gdist.inputs = gdist_inputs,
  GA.inputs = ga_inputs,
  dist_mod = FALSE,
  null_mod = FALSE,
  diagnostic_plots = FALSE
)
```

The optimization result is still a named list, but it now carries an S3 class
with package methods for `print()`, `summary()`, and `plot()`.

```{r ss-optim-s3, eval=FALSE}
class(ss_out)
summary(ss_out)
plot(ss_out)
```

## Multi-Surface Optimization

```{r ms-optim, eval=FALSE}
ms_out <- MS_optim(
  gdist.inputs = gdist_inputs,
  GA.inputs = ga_inputs,
  diagnostic_plots = FALSE
)
```

```{r ms-optim-s3, eval=FALSE}
class(ms_out)
summary(ms_out)
plot(ms_out)
plot(ms_out, type = "fit")
```

If you want to reconstruct the optimized composite surface after a multisurface run:

```{r combine-surfaces, eval=FALSE}
combined_surface <- Combine_Surfaces(
  PARM = ms_out$GA.summary@solution[1, ],
  gdist.inputs = gdist_inputs,
  GA.inputs = ga_inputs
)
```

## All-Combinations Analysis

For `all_comb()`, create the GA settings with `Results.dir = "all.comb"` and provide a separate directory where outputs should be written. That output directory will be created automatically when needed.

```{r all-comb, eval=FALSE}
ga_allcomb <- GA.prep(
  raster = raster_orig,
  Results.dir = "all.comb",
  monitor = FALSE,
  quiet = TRUE
)

results_dir <- file.path(tempdir(), "ResistanceGA2-all-comb")

combo_out <- all_comb(
  gdist.inputs = gdist_inputs,
  GA.inputs = ga_allcomb,
  results.dir = results_dir,
  max.combination = 2,
  iters = 10,
  replicate = 1
)
```

```{r all-comb-s3, eval=FALSE}
class(combo_out)
summary(combo_out)
plot(combo_out)
```

# Working with Transformations

The package still provides the transformation utilities used during
optimization. These are useful for inspecting the shape of hypothesized
relationships before running a full search.

The figure below recreates the eight non-flat transformation families using a
common illustrative parameter set (`shape = 2.5`, `max = 100`) across the
range of the example continuous surface. The goal is to show the geometry of
the available curve families, not to imply that these are the parameter values
that should be used in an analysis. The flat Distance transformation is not
shown because it simply assigns a constant resistance of 1 everywhere.

```{r transformation-families, echo=FALSE, fig.width=7, fig.height=12, fig.cap="*Illustrative transformation families available for continuous resistance surfaces. All panels use the same shape and maximum parameters to emphasize differences in curve geometry.*"}
scale_vector <- function(x, min_out, max_out) {
  ((x - min(x)) * (max_out - min_out) / (max(x) - min(x))) + min_out
}

transform_curve <- function(transformation, x, shape, max_value) {
  if (transformation == "Inverse-Reverse Monomolecular") {
    y <- -max_value * (1 - exp(-x / shape)) - 1
    rev(scale_vector(y, abs(max(y)), abs(min(y))))
  } else if (transformation == "Inverse-Reverse Ricker") {
    y <- -(max_value * x * exp(-x / shape)) - 1
    rev(scale_vector(y, abs(max(y)), abs(min(y))))
  } else if (transformation == "Monomolecular") {
    max_value * (1 - exp(-x / shape)) + 1
  } else if (transformation == "Ricker") {
    max_value * x * exp(-x / shape) + 1
  } else if (transformation == "Reverse Monomolecular") {
    y <- max_value * (1 - exp(-x / shape)) + 1
    rev(y)
  } else if (transformation == "Reverse Ricker") {
    y <- max_value * x * exp(-x / shape) + 1
    rev(y)
  } else if (transformation == "Inverse Monomolecular") {
    y <- -max_value * (1 - exp(-x / shape)) - 1
    scale_vector(y, abs(max(y)), abs(min(y)))
  } else if (transformation == "Inverse Ricker") {
    y <- -(max_value * x * exp(-x / shape)) - 1
    scale_vector(y, abs(max(y)), abs(min(y)))
  } else {
    stop("Unknown transformation: ", transformation)
  }
}

curve_names <- c(
  "Inverse-Reverse Monomolecular",
  "Inverse-Reverse Ricker",
  "Monomolecular",
  "Ricker",
  "Reverse Monomolecular",
  "Reverse Ricker",
  "Inverse Monomolecular",
  "Inverse Ricker"
)

curve_params <- list(shape = 2.5, max_value = 100)
range_stats <- terra::global(raster_orig[["cont_orig"]], c("min", "max"), na.rm = TRUE)
original_values <- seq(range_stats$min, range_stats$max, length.out = 200)
scaled_values <- scale_vector(original_values, 0, 10)

curve_data <- do.call(
  rbind,
  lapply(curve_names, function(name) {
    data.frame(
      original = original_values,
      transformed = transform_curve(
        transformation = name,
        x = scaled_values,
        shape = curve_params$shape,
        max_value = curve_params$max_value
      ),
      transformation = name
    )
  })
)

curve_data$transformation <- factor(curve_data$transformation, levels = curve_names)

ggplot2::ggplot(curve_data, ggplot2::aes(x = original, y = transformed)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::facet_wrap(~ transformation, ncol = 2, scales = "free_y") +
  ggplot2::theme_bw() +
  ggplot2::labs(
    x = "Original data values",
    y = "Transformed resistance values"
  ) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold", size = 9),
    axis.text = ggplot2::element_text(size = 9),
    axis.title = ggplot2::element_text(size = 10)
  )
```

Monomolecular families are usually the best default because they are easier to
interpret and generally less prone to overfitting. Ricker families are helpful
when there is a strong biological reason to expect a peak or threshold-like
response, but they should usually be explored more cautiously. During
optimization, very large shape values for Ricker-family transformations are
collapsed to the flat Distance solution, because those curves can otherwise
approximate monomolecular behavior while adding unnecessary flexibility.

```{r plot-trans}
Plot.trans(
  PARM = c(2.5, 100),
  Resistance = raster_orig[["cont_orig"]],
  transformation = "Monomolecular",
  marginal.plot = FALSE
)

```

You can also apply a transformation directly to a raster:

```{r resistance-tran}
transformed_surface <- Resistance.tran(
  transformation = "Monomolecular",
  shape = 2.5,
  max = 100,
  r = raster_orig[["cont_orig"]]
)

transformed_surface
```

# Current Circuitscape Runs

The current map shown below is precomputed and stored as package data so the
vignette can render quickly and reproducibly.

```{r current-map}
vignette_current_map
```

To reproduce either the pairwise resistance distances or the current map
directly from Julia:

```{r run-julia, eval=FALSE}
cs_result <- Run_CS.jl(
  jl.inputs = jl_inputs,
  r = resistance_surfaces[["continuous"]]
)

current_map <- Run_CS.jl(
  jl.inputs = jl_inputs,
  r = resistance_surfaces[["continuous"]],
  CurrentMap = TRUE,
  output = "raster",
  EXPORT.dir = results_dir
)
```

If you want direct `.ini`-style access to Circuitscape from R, use
`Run_CS.ini()` instead of `Run_CS.jl()`. That lower-level wrapper is described
in the `Direct Circuitscape Runs from R` vignette.

# Migration Notes

The original `ResistanceGA` package relied on workflows that are no longer part of `ResistanceGA2`.

- Do not pass directories of `.asc` files as user input.
- Do not pass `sp` classes such as `SpatialPoints`.
- Do not use `JuliaCall` or `XRJulia`.
- Do not use the standalone Circuitscape executable.
- Do not use the old smoothing workflow based on `scale = TRUE` or `k.smooth()`.

If you keep those changes in mind, the modern workflow is mostly a matter of preparing terra objects up front and choosing whether the distance engine should be `gdistance` or Julia/Circuitscape.
